---
title: "R getting started - session 1"
theme: "CambridgeUS"
colortheme: "seahorse"
author: 
- Alina Ferecatu \newline 
- (with contributions from Pieter Schoonees and Andreas Alfons) \newline
- Rotterdam School of Management \newline
- Erasmus University
date: "9/2/2020"
fontsize: 11pt
output: 
  beamer_presentation:
    
    slide_level: 2
    fig_width: 5
    fig_height: 4
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


## Documents and software

Have the **latest version** of:

- R: [https://CRAN.R-project.org](https://CRAN.R-project.org)
- RStudio Desktop: [https://www.rstudio.com](https://www.rstudio.com)

- Installation instructions on Canvas

## Today's lecture

Getting started with R:

- Overview of R ecosystem
- Load data
- Descriptive statistics
- Graphics

# Overview of R

## About R 

- Open source (free) environment for statistical computing
- One of the most popular data science tools worldwide
- Runs on Linux/Unix, Mac OS X, Microsoft Windows
- Fully developed and easy-to-use programming language
- Extensible by community contributed packages
    - 13500+ packages on CRAN, Bioconductor and Github

## What is R used for?

1. Collecting data
    - Scrape from the web, import from databases, ...
    
2. Preparing, exploring and cleaning data
    - Data wrangling, exploratory data analysis, plotting

3. Modelling
    - Regression, segmentation, machine learning, custom methods, ...

4. Model evaluation
   - Assessing model quality

5. Reporting results
  - Writing (dynamic) reports, visualization, creating interactive web applications, ...
  
And more (including everything SPSS can do)

## Who uses R
::: columns

:::: column

- Google
- Twitter
- Facebook
- New York Times
- John Deere
- Deloitte
- Credit Suisse
- Novartis
- eBay
::::

:::: column
 
 - Ford Motor Company
 - Kickstarter
 - Uber
 - Airbnb
 - Booking.com
 - Bank of America
 - McKinsey & Company
 - FourSquare
 - ...

::::

:::

**You too?**

## RStudio: four panels

- Top left: Script editor (if open)

- Bottom left: R console

- Top right: *Two tabs*

    - Environment: list of objects used in the session
    - History: allows to re-run previous commands 

- Bottom right: Five tabs

    - Files: browse through filles on the computer

    - Plots: graphics are displayed here

    - Packages: list of installed packages

    - Help: R help files are displayed here

    - Viewer: local web content created in the session

## Example session 1

Please open RStudio, and open example1.R in the script editor

      - Available on Canvas
      
      - Execute the line in which your cursor is 
      with Ctrl / Cmd + Enter
      
      - You can also type the command directly in the 
      console, but it is better to store your commands 
      separately in a script (reproducibility)

## Script files in RStudio

Create a new script file: File ! New file -> R Script

Save script file:

  - Keyboard shortcut: *Ctrl / Cmd + S*
  
  - File -> Save As. . .  and enter the file name in the dialog
  
---> Use file extension .R
 
Open existing script file:

  - File -> Open File. . . and select the script file in the dialog
  
  - Click the Files tab in the lower right panel, navigate to the script file
and click on it

## Some details

 - R is case sensitive
 
 - The + prompt means R is waiting for you to complete the command
 
 - Press Esc in the console to cancel the command being evaluated
 
 - Use the Tab key for code completion
 
 - Remember to close your parentheses ()
 
## Install packages

1. From CRAN

*Tools --> Install packages ...*

In the dialog:
  - In the Install from: box, select Repository 
  
  - Type the names of the packages into the text box (suggestions are shown as you type)
  
  - Make sure that Install dependencies is checked 
  
  - Click Install
  
2. From command line

``` {r, eval=FALSE}
install.packages("tidyverse")
```
 
## Load installed packages

In RStudio:

1. In the lower right panel, click the Packages tab

2. Check the box next to the packages to be loaded

On the command line:

``` {r } 
library("tidyverse")
```

--> Install a package **once** on a computer

--> Load it in every new session

# Data and descriptive statistics

## Loading data

R can read all sorts of data:
 - Native R data files: .RData or .rda
 
 - Text files: .txt and .csv
 
 - Excel spreadsheets: .xls(x)
 
 - SPSS files: .sav and .por
 
... and many others

--> We will use the simplest for now: RData format

--> RStudio makes importing other common data types easy: File --> Import Dataset

## Load RData files: RStudio

RStudio:
Click the Files tab in the lower right panel, navigate to the R data file and click on it

Or: File --> Open File. . . and select the R data file in the dialog

--> R objects loaded into your session environment (workspace)

## Load RData files: command line

Path relative to the current working directory:

``` {r, eval=FALSE}
load("Prestige.RData")
```
 
If the file is not in the working directory, specify the full path:

``` {r}
load("~/Documents/CMR/session_1/Prestige.RData")
```
 

Always use / and not \\ (Windows!)

## Working directory

R follows a one directory per project philosophy

--> Location where R starts looking for files on the file system

**In RStudio:**

1 Session -> Set Working Directory -> Choose Directory. . .

2 In the dialog, select the desired working directory

**From the command line:**

``` {r}
setwd("~/Documents/CMR/session_1")
```
  
You can automate this by using RStudio projects:
  
--> Opening the RStudio project restores the working directory

## View loaded data set objects

In RStudio:

1 In the top right panel, click the Environment tab

2 Click on the data set object name

On the command line:

--> Type the name of the data set to print it

--> Use the View() function to open RStudio's viewer

## View data sets: command line

Prestige of occupations in Canada in a data frame called Prestige:

--> Too much output even for moderately sized data sets

``` {r}
Prestige
```


## View first rows of data: head()

--> Get overview of what the data looks like

``` {r}
head(Prestige)
```

View last rows of the data:
``` {r, eval=F}
tail(Prestige)
```

## Summarize data: summary()
``` {r}
summary(Prestige)
```

## Basic Data Frame Functions

Dimensions using dim():
```{r}
dim(Prestige)
```

Number of rows and columns separately with nrow() and ncol():


```{r}
nrow(Prestige)
```

```{r}
ncol(Prestige)
```


Column (variable) names:
```{r}
colnames(Prestige)
```


## Extracting Variables
One way of extracting variables from a data.frame is via the $ operator:

```{r}
Prestige$education
```

--> Type Prestige$ and RStudio will bring up suggestions


## Basic Mathematical Functions

::: columns

:::: column

*Operator or function*

\+ 

\- 

\* 

/ 

^ 

abs() 

sqrt() 

log()

exp() 
::::

:::: column

*Operation*

addition x + y

subtraction x - y

multiplication x * y

division x / y

exponentiation x ^ y

absolute value abs(x)

square root sqrt(x)

logarithm log(x)

exponential function exp(x)

::::

:::


## Basic Math Example

*women* as a proportion:

```{r}
Prestige$women / 100
```

## Adding a New Variable

Assign a value to a new name, using $ with <-:
```{r}
Prestige$income <- 2 ^ Prestige$logincome
```

Check that it worked:
 
```{r}
head(Prestige)
```

## Basic Data Analysis Functions: Vectors


::: columns

:::: column

*Function* 

length(x)

min(x)    

max(x)    

sum(x)  

range(x) 

quantile(x) 

mean(x) 

var(x) 

sd(x)

median(x) 

table(x)

table(x, y)

cor(x, y) 

cov(x, y)

::::

:::: column

*Operation*

 Length
 
 Minimum
 
 Maximum
 
 Summation (total)
 
 Minimum and maximum
 
 Quantiles
 
 Mean
 
 Variance
 
 Standard deviation
 
 Median
 
 Contingency table
 
 Cross-tabulation
 
 Correlation
 
 Covariance
 
 
::::

:::

## Example

Minimum and maximum percentage of women:
```{r}
 min(Prestige$women)

max(Prestige$women)

range(Prestige$women)
```

Mean, median and standard deviation of logincome:

```{r}
mean(Prestige$logincome)

median(Prestige$logincome)
sd(Prestige$logincome)
```

## Functions

Frequency (contingency) table of type:
```{r}
table(Prestige$type)
```

## Missing values
You have to specify how R must handle missings (NA):
--> Depends on function; usually using the na.rm argument
```{r}
table(Prestige$type, useNA = "always")
```

--> See the help page, for example, ?table and ?mean
--> Default behaviour depends on function

Pearson correlation between women and prestige:

```{r}
cor(Prestige$women, Prestige$prestige) 
```


## Recap: Special Values

::: columns

:::: column
NA 

NaN 

Inf 

-Inf

NULL

pi

::::

:::: column

Not available (represents missing value)

Not a number (usually result of division 0/0)

Positive infinity

Negative infinity

Represents undefined value

 
::::

:::

## Practice

Consider the patents data in the file patents.RData

Information on patents granted in 2012 in each US state

Variables are described on the next slide

The data were scraped from StatsAmerica and Wikipedia

## Variables

total The total number of granted patents.

utility The number of granted utility patents.

design The number of granted design patents.

plant The number of granted plant patents.

population The number of inhabitants.

area Land area in km2.

governor Party afiliation of the state governor.

## 

area Land area divided into three categories: "small", "medium" and "large".

density The population density.

densitycat Population density divided into two categories: "low" and "high".

logdensity Logarithm of population density.

logutility log(utility+1).

logdesign log(design+1).

logplant log(plant+1).

## Exercises 1: Questions

1 What is the minimum and maximum number of total patents granted per state?

2 What is the Pearson correlation between logtotal and logdensity?

3 How many Republican governors were there in 2012? And how many Democrats? Use the governor variable.

4 How many Republican governors were there in small states? How many Democratic governors were there in large states? Use the areacat variable.

5 What is the mean proportion of the total patents per state that consist of utility patents? And the median and the standard deviation?


# Basic plotting with package ggplot2

## Graphics systems in R

R has several plotting systems and packages, including:

- Base graphics (e.g., plot()) is older but still very good

- Package ggplot2 delivers modern, cutting-edge capabilities:

--> We focus on ggplot2 graphics

--> Whatever analysis you do, always check if you can *plot()* the result

## Scatterplot matrix

```{r,  fig.height = 3, fig.width = 5}
plot(Prestige)
```


## The grammar of graphics

 - Implemented in package ggplot2

 - Designed with recent research on data visualization and human perception in mind

 - Focused on coherence between geometry of the data and geometry of the plot

 - The visual representation should fit the data

--> Must explicitly specify what variables to use and how to plot them

Ref: Wickham (2009): ggplot2: Elegant Graphics for Data Analysis

## Basic usage of ggplot2
Add together two basic elements:

1. Scaffolding deffned by ggplot()

    - Selects the data set
    
    - Deffnes the variables to be used (the aesthetic mapping): function aes()

2. Any number of visual representations of the data, known as geoms

    - Define the visual representation (the geometric objects): function family geom_x()
    
    - Different elements are added to the plot using the + operator
    
Load the package:

```{r,  fig.height = 2, fig.width = 3}
library("ggplot2")
```

## Scatterplot: Scaffolding

```{r,  fig.height = 2, fig.width = 3}
ggplot(Prestige, aes(x = prestige, y = logincome))
```


## Scatterplot: Scaffolding + points

```{r,  fig.height = 2, fig.width = 3}
ggplot(Prestige, aes(x = prestige, y = logincome)) +
  geom_point()
```

## Histogram

```{r,  fig.height = 2, fig.width = 3}
ggplot(Prestige, aes(x = prestige)) +
  geom_histogram()
```

## Histogram: number of bins

--> For histograms, it is always a good idea to play with the number of bins

  - Number of bins can be specified with argument bins
  
  - Bin width can be specified with argument binwidth

## Histogram: number of bins
```{r, fig.height = 2, fig.width = 3}
ggplot(Prestige, aes(x = prestige)) +
  geom_histogram(bins = 15)
```

## Density plot

```{r, fig.height = 2, fig.width = 3}
ggplot(Prestige, aes(x = prestige)) +
  geom_density()
```


# Extras 

## R help

R comes with built-in help for more information on functionality:

 - Help topic is usually the name of a function, data set or package

 - Help files are required for packages on CRAN
 
 - Overview of all help files within a package is available
 
In RStudio:

1 In the lower right panel, click the Help tab

2 Type the topic into the text box on the right (suggestions are shown as you type) 
 
## View R help files: command line

Help for the help() function is available:

``` {r, echo = TRUE} 
?help
help("help")
```

List all help topics within a package:
``` {r, echo = TRUE} 
help(package = "ggplot2")
help("help")
```
 
Run examples from a help file:

``` {r, echo = TRUE} 
example("mean")
```
