---
title: "R getting started - session 1"
theme: "CambridgeUS"
colortheme: "seahorse"
author: 
- Alina Ferecatu \newline 
- (with contributions from Pieter Schoonees and Andreas Alfons) \newline
- Rotterdam School of Management \newline
- Erasmus University
date: "9/2/2020"
output: 
  beamer_presentation:
    slide_level: 2
    fig_width: 7
    fig_height: 6
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Documents and software

Have the **latest version** of:

- R: [https://CRAN.R-project.org](https://CRAN.R-project.org)
- RStudio Desktop: [https://www.rstudio.com](https://www.rstudio.com)

- Installation instructions on Canvas

## Today's lecture

Getting started with R:

- Overview of R ecosystem
- Load data
- Descriptive statistics
- Graphics

# Overview of R

## About R 

- Open source (free) environment for statistical computing
- One of the most popular data science tools worldwide
- Runs on Linux/Unix, Mac OS X, Microsoft Windows
- Fully developed and easy-to-use programming language
- Extensible by community contributed packages
    - 13500+ packages on CRAN, Bioconductor and Github

## What is R used for?

1. Collecting data
    - Scrape from the web, import from databases, ...
    
2. Preparing, exploring and cleaning data
    - Data wrangling, exploratory data analysis, plotting

3. Modelling
    - Regression, segmentation, machine learning, custom methods, ...

4. Model evaluation
   - Assessing model quality

5. Reporting results
  - Writing (dynamic) reports, visualization, creating interactive web applications, ...
  
And more (including everything SPSS can do)

## Who uses R
::: columns

:::: column

 - Google
::::

:::: column
 
- NYT
::::

:::

**You too?**

## Slide with Plot

```{r pressure}
plot(pressure)
```

